\chapter{Mathematical Background}
\label{ch:background}

This chapter introduces the notation and mathematical foundations
used throughout the thesis.


% ==================================================================
\section{Finite Fields}
\label{sec:finite_fields}
% ==================================================================

\begin{definition}[Prime Field]
\label{def:prime_field}
Let $p$ be a prime number. The \emph{prime field} $\Fp$ is the set
$\{0, 1, \ldots, p-1\}$ equipped with addition and multiplication
modulo $p$.
\end{definition}

The multiplicative group $\Fp^* = \Fp \setminus \{0\}$ is cyclic of
order $p - 1$. An element $g \in \Fp^*$ is called a \emph{generator}
if $\{g^0, g^1, \ldots, g^{p-2}\} = \Fp^*$. For any $a \in \Fp^*$,
the multiplicative inverse can be computed via Fermat's little theorem:
%
\begin{equation}
\label{eq:fermat}
a^{-1} \equiv a^{p-2} \pmod{p}.
\end{equation}


\subsection{The BabyBear Field}
\label{subsec:babybear}

The implementation described in this thesis operates over the BabyBear
prime field, defined by
%
\begin{equation}
\label{eq:babybear_prime}
p = 2^{31} - 2^{27} + 1 = 2\,013\,265\,921.
\end{equation}

% p < 2^31 so elements fit in 32 bits; products fit in 64 bits.
Since $p < 2^{31}$, every field element fits in a 32-bit unsigned
integer, and the product of two elements fits in a 64-bit integer.

% p - 1 = 2^{27} * 15, giving a multiplicative subgroup of order 2^{27}.
The multiplicative group $\Fp^*$ has order $p - 1 = 2^{27} \cdot 15$.
The factor $2^{27}$, the \emph{two-adicity} of the field, means that
$\Fp^*$ contains a cyclic subgroup of order $2^{27}$, which is
essential for the Number Theoretic Transform.


\subsection{Roots of Unity}
\label{subsec:roots_of_unity}

\begin{definition}[Primitive Root of Unity]
\label{def:root_of_unity}
An element $\omega \in \Fp$ is a \emph{primitive $N$-th root of unity}
if $\omega^N = 1$ and $\omega^k \neq 1$ for all $0 < k < N$.
\end{definition}

Given a primitive $N$-th root of unity $\omega$, the set
$H = \{1, \omega, \omega^2, \ldots, \omega^{N-1}\}$
is a multiplicative subgroup of $\Fp^*$ of order $N$. Because BabyBear
has two-adicity $27$, subgroups of order $2^k$ for any $k \leq 27$
exist and are generated from a fixed multiplicative generator $g$ by
taking $g^{(p-1)/2^k}$.

% ethSTARK Section 3 (Definition 1, AIR): "the following definition
% restricts our attention only to finite fields F that contain a large
% multiplicative subgroup of size 2^h".
These subgroups serve as evaluation domains for the polynomials in the
STARK protocol: the execution trace is evaluated on a subgroup $H$ of
order $2^h$, and FRI operates on a larger domain.


\subsection{Field Extensions}
\label{subsec:field_ext}

\begin{definition}[Field Extension]
\label{def:field_ext}
A \emph{field extension} $\mathbb{F}_{p^k}$ of degree $k$ over $\Fp$
is the quotient ring $\Fp[X]/(f(X))$, where $f(X) \in \Fp[X]$ is an
irreducible polynomial of degree $k$. Elements are polynomials of
degree less than $k$ with coefficients in $\Fp$.
\end{definition}

In this work we use the \emph{quartic extension} $\Fpk$, constructed
via the irreducible polynomial $X^4 - 11$. The constant $11$ is chosen
because it is not a fourth power in $\Fp$, which guarantees that
$X^4 - 11$ is irreducible. An element of $\Fpk$ is a tuple
$(c_0, c_1, c_2, c_3) \in \Fp^4$ representing
$c_0 + c_1 X + c_2 X^2 + c_3 X^3$, with the reduction rule $X^4 = 11$.


% ==================================================================
\section{Polynomials}
\label{sec:polynomials}
% ==================================================================

A univariate polynomial over $\Fp$ of degree at most $d$ is
$f(X) = \sum_{i=0}^{d} c_i X^i$ with $c_i \in \Fp$. It can be
represented by its \emph{coefficient vector} $(c_0, \ldots, c_d)$ or
by its \emph{evaluation vector} on a domain $D$. Converting between
the two representations, and evaluating a polynomial at a point, are
the fundamental operations of the prover.


\subsection{The Number Theoretic Transform}
\label{subsec:ntt}

The conversion between coefficient and evaluation representations is
performed using the Number Theoretic Transform (NTT), the finite-field
analogue of the Fast Fourier Transform. When the evaluation domain is
a multiplicative subgroup
$D = \{1, \omega, \ldots, \omega^{N-1}\}$ with $N$ a power of~$2$, the
NTT computes all evaluations $f(\omega^i)$ in $O(N \log N)$ operations.
The algorithm exploits the decomposition
%
\[
f(X) = f_{\mathrm{even}}(X^2) + X \cdot f_{\mathrm{odd}}(X^2),
\]
%
recursing on the even- and odd-indexed coefficients over the squared
subgroup, which has half the size. The inverse transform runs the same
routine with $\omega^{-1}$ and divides by $N$.


\subsection{Low-Degree Extension}
\label{subsec:lde}

The \emph{low-degree extension} (LDE) of a polynomial is its
evaluation on a domain larger than its coefficient count. Given the
$n$ coefficients of $f$, the LDE with blowup factor $b$ evaluates $f$
on a coset of a subgroup of size $b \cdot n$. The resulting evaluation
vector is the Reed-Solomon codeword that the prover commits to. In our
implementation the low-degree extension is computed by padding the
coefficients with zeros, applying a coset shift, and running the NTT
on the enlarged domain.
$D \supseteq H$~\cite[Section~3.4]{ethSTARK}

\subsection{Vanishing Polynomials}
\label{subsec:vanishing}

% ethSTARK, before Definition 1 (AIR): "Given a set S in F, we define
% the vanishing polynomial of S to be Z_S(X) := prod_{s in S}(X - s)".
\begin{definition}[Vanishing Polynomial {\cite[Section~5.1]{ethSTARK}}]
\label{def:vanishing}
Given a set $S \subseteq \F$, the \emph{vanishing polynomial} of $S$
is $Z_S(X) := \prod_{s \in S}(X - s)$. When $S$ is a multiplicative
subgroup $H = \{1, \omega, \ldots, \omega^{N-1}\}$, this simplifies
to $Z_H(X) = X^N - 1$.
\end{definition}

$Z_H$ vanishes on every element of $H$ and nowhere else. If a
constraint polynomial $c(X)$ must be zero on $H$, then $Z_H(X)$
divides $c(X)$, and the quotient $q(X) = c(X)/Z_H(X)$ is a
polynomial. The verifier checks that $q$ has low degree, which
certifies that $c$ vanishes on $H$~\cite{ethSTARK}. This mechanism is
the basis of the STARK construction presented in
Chapter~\ref{ch:stark}.


% ==================================================================
\section{Reed-Solomon Codes}
\label{sec:reed_solomon}
% ==================================================================

% BS+18, Section 2.1: "For S in F and rate parameter rho in (0,1), the
% Reed-Solomon code RS[F, S, rho] is the family of functions f : S -> F
% that are evaluations of polynomials of degree < rho|S|."
\begin{definition}[Reed-Solomon Code {\cite[Section~2.1]{BBHR18}}]
\label{def:rs_code}
For a finite field $\F$, a subset $S \subseteq \F$, and a rate
parameter $\rho \in (0,1)$, the \emph{Reed-Solomon code}
$\RS[\F, S, \rho]$ is the set of functions $f : S \to \F$ that are
evaluations of polynomials of degree less than $\rho|S|$:
%
\[
\RS[\F, S, \rho] = \{(f(x))_{x \in S} : f \in \F[X],\;
  \deg(f) < \rho|S|\}.
\]
\end{definition}

Each codeword is the evaluation of a polynomial of degree less than
$\rho|S|$ on all points of $S$. The code has minimum distance
$|S| - \rho|S| + 1$, so two distinct codewords differ on a large
fraction of positions.

The parameter $\rho$ is the \emph{rate}, and $1/\rho$ is the
\emph{blowup factor}. In our implementation the blowup factor is~$2$
($\rho = 1/2$): the codeword produced by the low-degree extension is
twice the length of the coefficient vector.


\subsection{Relative Hamming Distance}
\label{subsec:hamming}

% BS+18 Section 2.1: "the case that f is 0.1-far from (all members of)
% RS[F, S, rho] in relative Hamming distance."
\begin{definition}[Relative Hamming Distance]
\label{def:hamming}
The \emph{relative Hamming distance} between two vectors
$u, v \in \F^n$ is
%
\[
\Delta(u, v) = \frac{|\{i : u_i \neq v_i\}|}{n}.
\]
%
The \emph{distance} of a vector $u$ from a code $C$ is
$\Delta(u, C) = \min_{v \in C} \Delta(u, v)$. A vector $u$ is
$\delta$-\emph{close} to $C$ if $\Delta(u, C) \leq \delta$, and
$\delta$-\emph{far} if $\Delta(u, C) > \delta$.
\end{definition}

The FRI protocol tests proximity to a Reed-Solomon codeword. As stated
in~\cite[Section~2.1]{BBHR18}, the verifier's task is to distinguish
between the case that $f \in \RS[\F, S, \rho]$ and the case that $f$
is far from $\RS[\F, S, \rho]$ in relative Hamming distance.


% ==================================================================
\section{Cryptographic Hash Functions}
\label{sec:hash_functions}
% ==================================================================

\begin{definition}[Collision-Resistant Hash Function]
\label{def:hash}
A \emph{collision-resistant hash function} is a function
$H : \{0,1\}^* \to \{0,1\}^\lambda$ such that no efficient algorithm
can find distinct inputs $x \neq x'$ with $H(x) = H(x')$, except
with negligible probability.
\end{definition}

Hash functions serve two roles in STARK systems: constructing Merkle
trees for committing to polynomial evaluations, and deriving the
verifier's challenges via the Fiat-Shamir heuristic. Our
implementation uses two different hash functions for these two roles:
SHA-256 for the Merkle commitments and Keccak-$f$[1600] for the
challenger.

% CO25 proves DSFS (Duplex Sponge Fiat-Shamir) preserves soundness,
% knowledge-soundness and zero-knowledge.
The distinction matters for the Fiat-Shamir transformation. Chiesa and
Orr\`{u}~\cite{CO25} prove that a \emph{duplex sponge} construction
(such as Keccak/SHA-3) is suitable for Fiat-Shamir, while
Merkle-Damg\r{a}rd constructions (such as SHA-256) lack a comparable
security proof in this setting. For this reason we use a Keccak-based
sponge for the challenger, while retaining SHA-256 for the Merkle
trees, where only collision resistance is required.


% ==================================================================
\section{Merkle Trees}
\label{sec:merkle_trees}
% ==================================================================

\begin{definition}[Merkle Tree]
\label{def:merkle}
A \emph{Merkle tree} over a sequence $(v_0, v_1, \ldots, v_{N-1})$
with $N = 2^k$ is a binary tree where each leaf stores the hash of one
element, $\ell_i = H(v_i)$; each internal node stores the hash of the
concatenation of its two children,
$H(n_{\mathrm{left}} \| n_{\mathrm{right}})$; and the root $r$ is a
single digest that binds the entire sequence.
\end{definition}

To prove that $v_i$ is the value at position $i$, the prover provides
an \emph{authentication path}: the $\log_2(N)$ sibling hashes along
the path from the leaf to the root. The verifier recomputes the root
and accepts if it matches $r$. Verification requires $O(\log N)$ hash
evaluations.

In the BCS transformation~\cite{BCS16}, Merkle trees commit to the
oracle messages of the IOP: the prover evaluates a polynomial on the
FRI domain, builds a Merkle tree over the evaluations, and sends the
root to the verifier.


% ==================================================================
\section{Proof Systems}
\label{sec:proofs}
% ==================================================================

We now establish the formal framework of proof systems, following
Thaler~\cite{Thaler22}.


\subsection{Interactive Proofs}
\label{subsec:interactive_proofs}

% Thaler, Definition 3.1.
\begin{definition}[Interactive Proof {\cite[Definition~3.1]{Thaler22}}]
\label{def:interactive_proof}
An \emph{interactive proof system} for a function $f$ is a pair
$(\Prover, \Verifier)$ where $\Prover$ is a computationally unbounded
prover and $\Verifier$ is a probabilistic polynomial-time verifier.
The protocol proceeds in rounds, and must satisfy:
\begin{itemize}
  \item \textbf{Completeness.} For every input $x$,
    $\Pr[\Verifier \text{ accepts}] \geq 1 - \delta_c$.
  \item \textbf{Soundness.} For every input $x$ and every cheating
    prover $\Prover^*$ that sends a value $y \neq f(x)$,
    $\Pr[\Verifier \text{ accepts}] \leq \delta_s$.
\end{itemize}
\end{definition}


\subsection{Arguments}
\label{subsec:arguments}

% Thaler, Definition 3.2.
\begin{definition}[Argument System {\cite[Definition~3.2]{Thaler22}}]
\label{def:argument}
An \emph{argument system} for a function $f$ is an interactive proof
for $f$ in which the soundness condition is only required to hold
against prover strategies that run in polynomial time.
\end{definition}

STARKs are \emph{arguments}, not \emph{proofs} in the formal sense:
their soundness relies on the collision resistance of the hash
function, which a computationally unbounded prover could break.


\subsection{The Fiat-Shamir Transformation}
\label{subsec:fiat_shamir}

A \emph{non-interactive} proof consists of a single message from the
prover to the verifier. The Fiat-Shamir transformation removes
interaction from a public-coin protocol by replacing the verifier's
random challenges with the output of a hash function applied to the
transcript so far. Under the random oracle model, this preserves
soundness. Chiesa and Orr\`{u}~\cite{CO25} give a formal treatment
using a duplex sponge, which resolves the ambiguity between writing to
and reading from the transcript by using the sponge's native
\emph{absorb} and \emph{squeeze} operations. Our challenger implements
this construction over Keccak-$f$[1600].


% ==================================================================
\section{Interactive Oracle Proofs}
\label{sec:iop}
% ==================================================================

% BCS16, Definition 2.2.
\begin{definition}[Interactive Oracle Proof
  {\cite[Definition~2.2]{BCS16}}]
\label{def:iop}
An \emph{Interactive Oracle Proof} (IOP) for a language $\mathcal{L}$
is an interactive protocol between a prover $\Prover$ and a verifier
$\Verifier$ that proceeds in rounds. In each round:
\begin{enumerate}
  \item the prover sends an \emph{oracle message} $\pi_i$ (a function
    that the verifier may query at chosen positions);
  \item the verifier sends a random challenge $r_i$.
\end{enumerate}
After $k$ rounds, the verifier makes a bounded number of queries to
$\pi_1, \ldots, \pi_k$ and outputs $\mathsf{accept}$ or
$\mathsf{reject}$.
\end{definition}

The key difference from a standard interactive proof is that the
verifier has \emph{oracle access} to the prover's messages: it reads
only a small number of positions rather than the full message. This
is what enables succinct verification.

An IOP of \emph{proximity} (IOPP) is an IOP in which the verifier
tests whether the prover's initial oracle is close to a codeword of
a specified code, rather than testing membership in a 
language~\cite{BBHR18}.


% ==================================================================
\section{The BCS Transformation}
\label{sec:bcs}
% ==================================================================

The BCS transformation~\cite{CO25} compiles a public-coin IOP into a
non-interactive argument in the random oracle model:

\begin{enumerate}
  \item \textbf{Commitment.} Each oracle message is committed using a
    Merkle tree; the prover sends the root.
  \item \textbf{Challenge derivation.} Each challenge is derived from
    the transcript via Fiat-Shamir.
  \item \textbf{Query answering.} The prover opens the requested
    positions together with their Merkle authentication paths.
  \item \textbf{Verification.} The verifier checks the paths,
    recomputes the challenges, and runs the IOP decision procedure.
\end{enumerate}

This is precisely the structure of our FRI commit phase: each folding
layer is committed with a Merkle tree, its root is absorbed into the
challenger, and the next folding challenge is squeezed from it. The
query and verification steps are performed when the verifier opens the
Merkle paths and checks folding consistency.


% ==================================================================
\section{Polynomial Commitment Schemes}
\label{sec:pcs}
% ==================================================================


\begin{definition}[Polynomial Commitment Scheme]
\label{def:pcs}
A \emph{Polynomial Commitment Scheme} (PCS) consists of four
algorithms:
\begin{itemize}
  \item $\Setup(\lambda, d) \to \mathsf{pp}$: generates public
    parameters for polynomials of degree at most $d$.
  \item $\Commit(\mathsf{pp}, f) \to C$: produces a commitment $C$
    to the polynomial $f$.
  \item $\Open(\mathsf{pp}, f, z) \to (v, \pi)$: produces a proof
    $\pi$ that $f(z) = v$.
  \item $\Verify(\mathsf{pp}, C, z, v, \pi) \to \{0,1\}$: checks
    the evaluation proof.
\end{itemize}
The scheme must satisfy correctness and binding.


The STARK system implemented in this thesis uses a \emph{FRI-based
PCS}~\cite{BBHR18}: a polynomial is committed via a Merkle tree over
its low-degree extension, and evaluation proofs are constructed
through the FRI low-degree testing protocol. This approach requires no
trusted setup (transparent) and relies only on hash functions
(post-quantum secure). The FRI protocol is described in detail in
Chapter~\ref{ch:fri}.
\end{definition}